Interactively Visualizing Multivariate Market Segmentation Using the R Package Lionfish

### 2024-10-01 manuscript.tex
script for explaining the heatmaps
rewrite of normalization section

%% almost as usual
\author{Matthias Medl\,\orcidlink{0000-0002-3354-4579}\\ Institute of Statistics \\ BOKU University \\ Vienna \And 
        Dianne Cook\,\orcidlink{0000-0002-3813-7155}\\ Econometrics and Business Statistics \\ Monash University \\ Melbourne \And
        Ursula Laa\,\orcidlink{0000-0002-0249-6439}\\ Institute of Statistics \\ BOKU University \\ Vienna}
%% \AND can be used instead of \And and starts a new line
\title{Interactively Visualizing Clustering in Multivariate Data Using lionfish}

@@

\begin{figure*}[h]
\centerline{\includegraphics[width=1\textwidth]{images/intro1.png}}
\caption{Examples of how $k$-means partitions 2D data with different association structure. If the correlation is high (c), the partitioning happens along the primary direction of the association.}
\label{kmeans-partition}
\end{figure*}

@@

The aim of this work is to build an interface that allows for an interactive exploration of partitioning solutions of multivariate data. The interface should enable

\begin{itemize}
\item visualizing a clustering solution with different tour methods, including manual changes to the viewing direction (manual tour)
\item brushing of (groups of) points for manual refinement of the cluster solution (spin-and-brush capabilities)
\item additional linked displays for comprehensive exploration of the solution
\item ability to update the displays based on user selections
\end{itemize}

@@

\subsection{Feature selection based on fractional analysis}

A first step may look into what features are important for the cluster solution. The interface provides different capabilities that support feature selection. First, a 2D tour view can be used to understand the sensitivity of the grouping to individual features, in particular when using the manual tour to remove one feature at a time. This is in particular useful when the starting view was obtained through a projection pursuit guided tour~\citep{CBCH94} that optimized for the separation between the labeled groups~\citep{lckl2005}.

In our applications we are considering market segmentation analysis on binary features. For this setting we have developed a heatmap display that we have found to be particularly useful in feature selection. The idea is to understand differences in the binary ratings between clusters, where we use different normalizations to better explore the data. For \(c_{ij}\) the count (number of true values in the binary feature) in the \(i\)-th cluster and \(j\)-th feature, we define

\[
\text{Total fraction}\ f_{ij}^{tot} = \frac{c_{ij}}{n}
\]
with normalization to the total number of observations $n$ for a general overview,

\[
\text{Intra cluster fraction}\  f_{ij}^{clust} = \frac{c_{ij}}{n_{i}}
\]
showing the counts relative to the cluster's size \(n_i\), to see the fraction of observations in cluster $i$ for which feature $j$ was true, and

\[
\text{Intra feature fraction}\ f_{ij}^{feat} = \frac{c_{ij}}{c_{j}}
\]
to see how the true values for feature $j$ are distributed across the different clusters relative, with normalization to the overall count of that feature \(c_j = \sum_i c_{ij}\).

\textcolor{red}{different symbol for each? indicat on the index value somehow?}

>>>>

%% almost as usual
\author{Matthias Medl\,\orcidlink{0000-0002-3354-4579}\\ Institute of Statistics \\ BOKU University \\ Vienna \And 
        Dianne Cook\,\orcidlink{0000-0002-3813-7155}\\ Econometrics and Business Statistics \\ Monash University \\ Melbourne \And
        Ursula Laa\,\orcidlink{0000-0002-0249-6439}\\ Institute of Statistics \\ BOKU University \\ Vienna}
%% \AND can be used instead of \And and starts a new line
\title{Interactively Visualizing Multivariate Market Segmentation Using the R Package Lionfish}

@@

\begin{figure*}[ht]
\centerline{\includegraphics[width=1\textwidth]{images/intro1.png}}
\caption{Examples of how $k$-means partitions 2D data with different association structure. If the correlation is high (c), the partitioning happens along the primary direction of the association.}
\label{kmeans-partition}
\end{figure*}

@@

The aim of this work is to build an interface that allows for an interactive exploration of partitions generated by clustering of multivariate market data. In some combinations of the features, one should be able to the separations as can be seen in the 2D example in Figure \ref{kmeans-partition}. The objectives for the interface are to enable:

\begin{enumerate}
\item Visualizing the partitions between clusters in combinations of features using tour methods, with both a grand tour and manual tour.
\item Interactively brushing groups of points to refine the cluster solution, like done with spin-and-brush tools.
\item Linking multiple displays to focus on particular clusters, and simplify the problem to better understand the clustering solution. 
\item Update displays based on user selections such as feature selection or cluster selection, or re-scaling.
\end{enumerate}

@@

\subsection{Feature and cluster relationships}

A first step is to assess which features are important for the cluster solution. The interface provides different capabilities that support feature selection: tour view and summary heatmaps.

The 2D tour view can be used to understand the sensitivity of the grouping to individual features. When using the manual tour to remove one feature at a time, the relationship between this variable and the partition between clusters can be examined. The ideal starting view is that obtained through a projection pursuit guided tour~\citep{CBCH94} that optimized for the separation between the labeled groups~\citep{lckl2005}.

The summary heatmaps provide an overview of the cluster compositions relative to the features. Consider the matrix,

\[
C = \left[ \begin{array}{cccc} c_{11} & c_{12} & \dots & c_{1p} \\
                                              c_{21} & c_{22} & \dots & c_{2p} \\
                                              \vdots &  \vdots & \ddots &  \vdots \\
                                              c_{k1} & c_{k2} & \dots & c_{kp}
                                              \end{array} \right]
\]

where $c_{ij}, i=1, ..., p$ (number of features); $j=1, ..., k$ (number of clusters) are a summary of each feature in each cluster. Because our examples use binary features,  the $c_{ij}$ is the number of 1's of feature $j$ in cluster $i$. %For this setting we have developed a heatmap display that we have found to be particularly useful in feature selection. The idea is to understand differences in the binary ratings between clusters, where we use different normalizations to better explore the data. For \(c_{ij}\) the count (number of true values in the binary feature) in the \(i\)-th cluster and \(j\)-th feature, we define
There are several ways that these values can be normalized to examine different aspects: $f_{ij}^{o} = \frac{c_{ij}}{n}$, $f_{ij}^{c} = \frac{c_{ij}}{n_{i}}$ and $f_{ij}^{f} = \frac{c_{ij}}{n_{j}}$. The first is the overall fraction, where counts are normalised by the overall number of observations $n$. It is generally not particularly useful. The second, $f^c$ is normalized relative to the size of each cluster, and is useful for comparing the features which contribute most to each cluster. The last, $f_f$ is normalized relative to each feature, so it is useful to examine how features are distributed across clusters \textcolor{red}{Not sure this last explanation is good - at least I don't understand what I've written}

%\[
%\text{Overall fraction}\ f_{ij}^{o} = \frac{c_{ij}}{n}
%\]
%with normalization to the total number of observations $n$ for a general overview,

%\[
%\text{Intra cluster fraction}\  f_{ij}^{clust} = \frac{c_{ij}}{n_{i}}
%\]
%showing the counts relative to the cluster's size \(n_i\), to see the fraction of observations in cluster $i$ for which feature $j$ was true, and

%\[
%\text{Intra feature fraction}\ f_{ij}^{feat} = \frac{c_{ij}}{c_{j}}
%\]
%to see how the true values for feature $j$ are distributed across the different clusters relative, with normalization to the overall count of that feature \(c_j = \sum_i c_{ij}\).
